Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from the North of the Swiss Alps

### 2022-05-28 manuscript/paper_shorter.Rmd
added shorter 5000 word version

  Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density. Using the  area North of the main ridge of the Swiss Alps in prehistoric times (7000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone. The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research. Our prediction of population development of the case study area accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed probability distribution of radiocarbon dates.

@@

  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (7000–1000 BCE)

@@

Prehistorians have long recognised demography as a fundamental force in human cultural evolution [@childe_man_1936]. However, despite decades of interest in the demographic dynamics of prehistoric societies, concrete estimates of population size and population density before written records remain elusive. Though the archaeological record provides multiple possible demographic proxies [@hassan_demographic_1981], lack of access to this data, and a lack of methodological tools for turning it into quantitative estimates, has often left the conclusions drawn from it vague and superficial [@hassan_demographic_1981]. As a result, 'expert estimates' transferred from ethnographic parallels have often taken the place of direct inference from archaeological evidence [@morris_measure_2013, @turchin_seshat_2015].

Early attempts to estimate the size and density of prehistoric populations from archaeological remains were made by Hack [-@hack_changing_1942], Colton [-@colton_prehistoric_1949], Frankfort [-@frankfort_town_1950] and Cook [-@cook_reconsideration_1946]. Cook and Heizer [-@cook_quantitative_1965, -@cook_relationships_1968] and Naroll [-@naroll_floor_1962] provided rules derived from ethnographic contexts for estimating the size of prehistoric populations. Cook and Heizer (1966) also made the first attempt to estimate the population growth rate during the Neolithic [@hassan_demographic_1981]. These approaches became increasingly common after the 1960s, with the emergence of processual archaeology and its focus on human–environment relations and population ecology [@hassan_demographic_1981].

Following a spell out of fashion with the post-processual turn, prehistoric demography has experienced a revival in recent years as part of the cultural evolution paradigm [@shennan_population_2000; @riede_climate_2009 and others in same issue]. Part of this resurgence can be explained by a renewed interest in human–environment relations and human impact on the natural environment, which necessarily requires an assessment of population size. Kintigh et al. [-@kintigh_grand_2014], for example, list human influence, dominance, population size, and population growth amongst their 'grand challenges' for archaeology in the 21st century.

But methodological advances have certainly also played a role. In the last decade, the 'dates as data' technique [@rick_dates_1987], which uses the frequency of radiocarbon dates as a proxy for population dynamics, was taken up and significantly developed in a series of publications from the UCL group [e.g. @shennan_regional_2013], and has since been widely applied to archaeological contexts around the world [@crema_review_2022].

The 'dates as data' or summed radiocarbon approach has contributed greatly to our understanding of prehistoric demography, but it is not without its critics [@attenbrow_hiscock_2015, @carleton_groucutt_2020]. While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to certain fundamental problems common to all approaches that rely on a single proxy [@french_manifesto_2021, @schmidt_approaching_2021]. We believe that these problems cannot be overcome by methodological refinements in this area alone. Instead, the Bayesian approach we develop in this paper offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates.

@@

Müller and Diachenko [-@muller_tracing_2019] have compiled a list proxies currently used for the estimation of population size in prehistory. These can be roughly divided into three groups: ethnographic analogies; deductive estimates from ecological/economic factors; and the interpolation of frequencies of archaeological features (e.g. settlements, structures, individual finds). Three basic problems are common to all these approaches:

1.  **Reliance on single proxy**: Most investigations are using only one source of evidence. Although multi-proxy approaches exist, in such cases the individual proxies only serve to support each other or the proxy used as the main estimator. There is no explicit combination of results to reduce the inaccuracy and biases inherent to single proxies.
2.  **Uncertainty in measurements**: All archaeological evidence comes with an inherent uncertainty which is carried through to derived measurements. However, in most studies of prehistoric population, individual curves are presented as estimates, and the potential error associated with these estimates is almost never specified. 
3.  **Lack of a transfer function**: By 'transfer function', we mean something that allows for the proxy data to be interpreted in terms of actual population size or density. This could be absolute, i.e. a numerical estimate of population, or relative, i.e. a means of scaling changes in the proxy value to changes in population. Lack of a suitable frameworks and 'calibration' data means that this is rarely presented alongside proxy estimates. In the best case scenario, there will be a qualitative assessment of the informative value of the proxy, but this does not sufficiently account for the complex nature of archaeological data.

Furthermore, the types of archaeological data commonly used as population proxies share a number of problematic characteristics. They are:

-   **Limited**: We have only incomplete data that can be used for these purposes, and they are usually not very informative.
-   **Unevenly distributed**: For example, although there is a good data on settlement frequencies for some regions, and these are sometimes have a very high temporal resolution, these regions are very unevenly distributed over time and space.
-   **Noisy**: Frequently individual proxies are strongly influenced by factors unrelated to population, for example taphonomic conditions or depositional biases.
-   **Unreliable**: Research strategies, research history and varying levels of resources available to researchers strongly affect the nature of compiled datasets. As a result, systematic distortions are the rule rather than the exception.
-   **Heterogeneous**: A wide range of proxies can potentially inform us about population dynamics, all with different spatio-temporal scales, granularity, information value, scales, and data formats.
-   **Indirect**: We will never have direct data on prehistoric population; only proxy data that is thought to be a reasonable substitute for it. The transfer functions that link the proxy data with the desired quantity (population) are unknown. At best, they may be estimated by comparison with historical examples or the (ethnographic) present. However, since living conditions, circumstances and economic parameters differ greatly between the recent history and prehistory, these transfer functions can only be considered a rough first approximation.
-   **Contradictory**: When considering several proxies, differences in transfer functions, data quality and noisiness inevitably lead to different results. The most common way to deal with this is to reject certain proxies for a certain time periods, but this is not performed on a consistent or quantitative basis.

Here we argue that many, if not all, of these problems can be ameliorated through a) the explicit, quantitative integration of multiple proxies; and b) the use of a Bayesian approach to take account of and estimate uncertainty.

@@

Many of the problems with archaeological population proxies are also encountered in contemporary demography. In response, demographers have increasing turned to Bayesian methods to estimate and forecast contemporary population dynamics. For example, Bryant and Zhang [-@bryant2018] consider Bayesian data modelling a solution to exactly the kind of problems that affect archaeological data. Specifically, Bayesian approaches are well suited for limited, unreliable and noisy data.  Various sources of data, even contradictory data, can be brought into a common framework and used to support one another. These methods also provide a quantitative estimate of the likelihood and uncertainty of the model's resulting predictions (or in our case retro-dictions). Bayesian approaches are also capable of accounting for spatially and temporally incomplete data: where this data is missing, the uncertainty automatically increases, but this does not prevent general modelling and estimation. Finally, hierarchically-structured model suites, in which sub-models are created for each individual proxy, can be used to estimate transfer functions between individual proxies and the value to be modelled, thanks to the interaction of a large number of data sources and evidence.

This modeling technique can thus be used to combine different lines of evidence horizontally and vertically and in this way combine their results into a common conclusion and estimate, which at the same time includes an estimation of their reliability: if the data contradict each other, the overall reliability will be lower. If they support each other, the confidence interval will be smaller. And if there is no systematic bias that affects all data sources to the same extent, it should be possible to arrive at the most reliable estimate possible through the most heterogeneous set of data sources.

The well-established application of a Bayesian approach in radiocarbon calibration, where it is used to model radiometric uncertainty based on stratigraphic information, represents a very similar use case in archaeology. More recently, archaeologists have also used Bayesian modelling techniques as a tool for testing hypotheses relating to demographic trends or underlying models based on ^14^C data [e.g. @crema2021a]. This approach differs from the one presented in this paper in that, in these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only. This is a clear step forward in the direction of a model-based, and thus scientific, analysis. However, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of the inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool. Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of the time series. Thus, the credibility of a model can only be checked as a whole, without the dynamic developments that can arise in the course of demographic processes being represented. We would like to better exploit the capabilities of Bayesian hierarchical models through a combination of inductive data analysis and model-passed data integration of different proxies.

Here, we attempt to make Bayesian hierarchical techniques usable for archaeological reconstructions. In addition to a presentation of the basics and possible procedures, we want to show in the following, in a reproducible and practical form using a case study, how Bayesian methods can also make a decisive contribution to a better assessment of population development. These assessments are crucial for the reconstruction of the human past, even in for periods for which we only have very patchy, noisy and unreliable data.

@@

The basis of Bayesian statistics is the premise that there is always some prior assumption about the probability of an event, even if it may be very rough. This assumption is then adjusted by observing the data. The concept of probability is not derived from theoretical, infinitely repeatable random experiments or distributions, but from the direct confrontation of the pre-assumptions (priors) with the available data. This involves checking how credible these prior assumptions are with regard to the available data [likelihood, see also @bryant2018, 66]. A small amount of data leads to a broad probability distribution that is not strongly localised and restricted. The intuitive procedure is thereby the shifting of credibility through evidence.

At the heart of all Bayesian statistics is therefore the concept of updating a given prior assumption with new data and expressing this in probabilities [cf. also @kruschke2015, especially 15--25]. Our assumptions about the demographic development of the past must naturally be very conservative. In the logic of the basic Bayesian equation, these assumptions represent the prior (probability). In conjunction with the data that are included as the likelihood of the prior, a posterior (probability) results, which represents the Bayesian model learning from data. It is also in the nature of the approach that in real applications there is no point prediction, but in most cases a probability distribution for the prediction. Thus, we simultaneously obtained a result and an estimate of the confidence intervals, or rather the credibility interval, given the data.

This Bayesian learning is iterative and sequential, so that the result of one Bayesian inference can form the prior of another, i.e. it is an additive process [@kruschke2015, 17]. Moreover, at the conceptual level, this allows different sources of information to be combined [@bryant2018, 219--224]. This enables that they can be mapped to the same set and the same real-world domain, e.g. probabilities. This fact has long been exploited by archaeology in using stratigraphic information to make radiometric dating more accurate [@ramsey1995]. ^14^C dates and stratigraphy are something completely different, but both can be mapped to the probability of older and younger and combined in this way. The same is also feasible when it comes to the probability of population sizes or population densities or their derived dynamics.

A further characteristic of Bayesian modeling is that, due to the fact that results of a Bayesian inference can be regarded as a prior for the next one, a hierarchical formulation of problem domains is possible. Parameters that are necessary for an estimation, such as the relationship of population density to the deforestation signal in pollen data, need not be specified explicitly, but can be given by probability distributions and then estimated in the model itself [@bryant2018, 186]. The more data available, the more degrees of freedom can be estimated with a reasonable loss of confidence or a reasonable width of credibility intervals [@kruschke2015, 112]. For the estimation of these parameters, in turn, submodels have to be created which describe the relationship of the data to the characteristics of the parameter. This can be carried out over several levels, depending on necessity [@kruschke2015, 221--222].

@@

The case study area north of the Swiss Alps (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland’s territory and comprises the partly flat, but largely hilly area, between the Jura Mountains and the Alps. It averages an altitude of between 400 and 600 m a.s.l. It is a favourable area for settlement and agriculture and especially the Swiss Plateau is by far the most densely populated region of the Switzerland today. The Swiss Plateau between Lake Zurich and Lake Geneva will serve as a core region in our case study, because archaeological data from there is most abundant and accessible. Although it is a basin, depending on the region it has a very diverse natural landscape: shaped by glaciers during the ice ages, the many lakes and bogs that characterise the landscape today also provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, as well as a rich pool of sources for vegetation reconstructions by means of pollen analyses. Thanks to the very active and efficient archaeological research and heritage management there is an abundance of archaeological information, including known sites as well as dendrochronological and ^14^C data.

@@

Our case study targets the period between 7000–1000 BCE. The lower limit of this window was chosen to avoid the so-called 'Hallstatt plateau' on the Northern Hemisphere radiocarbon calibration curve, which would cause difficulties for the ^14^C proxy data. The upper limit was chosen to coincide with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence before 7000 in the working area.

In principle, a large number of different data sources can be integrated into the overall model as proxies, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created. Table \@ref(tab:tableproxies) provides a non-exhaustive list. For our case study, we used: a landscape openness indicator; an aoristic sum of sites based on typological dating; a sum calibration; and frequency data for dendro-dated lakeshore settlements in the Three Lakes region (western Swiss Plateau).

@@

With the vast areas of arable soils with small slope gradient compared to the Jura and the Alps, and a dense network of water bodies, the Swiss Plateau has always been a particularly favourable area for human settlement in Switzerland. From the Neolithic onwards, settlement areas were concentrated along its rivers and lakes [@christianlüthi2009]. Thus, our working region offers on the one hand excellent data for demographic estimation, but on the other hand poses very specific problems for such an undertaking. If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also might cause a research problem with regard to the ^14^C data often used as a proxy.

@@

The dataset we use for the number of dendro-dated wetland settlements in the Three Lakes region was collected by Julian Laabs for his PhD thesis [@laabs2019]. Details on the creation of this data series will be published at the referenced location. The time series used here runs from `r abs(min(dendro$date))` to `r abs(max(dendro$date))` BCE, and contains the number of chronologically registered fell phases at individual settlements. This results in a time series that reflects the settlement of the lakeshores in the Neolithic and Bronze Age periods.

@@

The dataset for the ^14^C sum calibration primarily consists of data from the XRONOS database (https://xronos.ch), supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of [@martínez-grau2021]. It contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites. The dates were selected so that their distribution area coincides with the catchment area of the pollen proxy (see also Figure \@ref(fig:c14map)). The dates in the dataset range in ^14^C years from `r max(c14$bp)` to `r min(c14$bp)` uncal BP. This time window extends beyond the study horizon in order to minimise boundary effects.

We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites. For the creation of the cumulative calibration, the corresponding functions of the R package rcarbon [@crema2021] were used with their default settings.

@@

```{r compdendroc14, fig.cap="Comparison of the scaled (non-centered z-transformed) number of ^14^C and dendro-dated sites over time in the dataset used."}
joined_data <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "preprocessed_data", "all_proxies.csv"))

ggplot(joined_data) + geom_line(aes(x=(1950-age)*-1, y=scale(sumcal, center = F), color = "darkred"), alpha = 0.75)  + geom_line(aes(x=(1950-age)*-1, y=scale(dendro, center = F), color = "darkblue"), alpha = 0.75) +
  theme_minimal() + scale_x_reverse() +
    labs(x = "years BCE",
         y = "scaled values",
         color = "Legend") + 
    scale_color_manual(values = c("darkblue", "darkred"), labels = c("dendro dated", "14C dated"))+
  theme(legend.position="bottom")
```

We can now compare these two data sets (Figure \@ref(fig:compdendroc14)). In fact, there is a not uninteresting fit between the two data series. However, it must be assumed that the two dating methods, even if they would contradict each other, actually complement each other, and thus allow a better overall unified picture of the actual settlement density than each of the individual proxies would allow on their own.

@@

To add another archaeological indicator of occupation, we include relative dating information obtained from the Swiss cantonal archaeology/heritage management authorities (Figure \@ref(fig:aoristmap)). These are primarily derived from scattered surface finds, which often have a low dating accuracy. We incorporate this data into our model as a typologically dated aorist time series. The dating accuracy is only in the range of archaeological periods, but the advantage is that we are not bound to the conditions and problems of radiocarbon dating and thus methodological issues of sum calibration can be avoided. Furthermore, this data provides an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by similar transmission filters and archaeological conditions as the evaluation of ^14^C data. Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy offered by archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.

```{r aoristcurve, fig.cap="Aoristic sum of the archaeological sites used in this analysis."}
ggplot(joined_data) + geom_line(aes(x=(1950-age)*-1, y=aoristic_sum, color = "darkgray"), alpha = 0.75) +
  theme_minimal()  +  scale_x_reverse() +
    labs(x = "years BCE",
         y = "aoristic sum",
         color = "Legend") + 
    scale_color_manual(values = "darkgray", labels = "Aoristic Sum")+
  theme(legend.position="bottom")
```

@@

The natural conditions of the lakes of the Swiss Plateau enable not only highly precise dating of archaeological sites, but also a very dense network of pollen analysis. We make use of this fact by generating a supra-regional openness indicator for the vegetation from the pollen data (Figure \@ref(fig:pollensites)). This proxy has the specific advantage that it is not dependent on preservation conditions, as archaeological indicators are. This makes it particularly valuable for indicating or compensating for systematic distortions that result from temporally specific settlement patterns and archaeological preservation conditions.

@@

The utilisation of this proxy is based on the assumption that the higher the population density in an area, the greater the human influence on the natural environment [@lechterbeck2014], and that the extensiveness of agricultural activity in an area is related to human population density [@zimmermann2004]. Evidence of land clearance in pollen diagrams can therefore provide further indications of population dynamics where humans can be assumed to be the main driver of this process, which is the case in much of Europe. The full procedure for deriving this proxy from several different pollen diagrams is detailed in a previous publication [@heitz2021]. In this study, five pollen diagrams from sites mainly in the hinterland of the large Alpine lakes were used. The technical steps are also documented in the accompanying R compendium. The percentage pollen data, based on a pollen sum of all terrestrial taxa of the individual sites, was combined into one data set by means of a principal component analysis (Figure \@ref(fig:pollenproxy)). Only terrestrial pollen taxa with a frequency of more than one third and, if present, with an average frequency of at least 0.1% were selected to reduce potential disturbance by rare species. Cereal pollen was explicitly retained as an important anthropogenic indicator. As each sample is absolutely dated, the data on the x-axis can be plotted against the openness value on the y-axis to obtain a time series time series for land clearance.

@@

## Other proxies

Unfortunately, there is no burial data available in the case study area that could be usefully incorporated into such this model due to the absence of regular burials for larger parts of prehistory. Formation conditions, in this case most likely C-transformation [@schiffer1987], the active cultural choice to not use archaeological traceable burial pratices, prevent their use in this study. Nevertheless, we see a very high potential for other regions in the integration of demographic indicators from burial data in order to enlarge the canon of methods and the range of proxies.

@@

The sum calibration, openness and the dendro-dated settlement data was smoothed by means of a moving average with a window of 50 years. The aoristic sum was not smoothed, because it already has a very coarse temporal resolution. The range of the smoothing window corresponded to the sample interval, with which a unified resolution of 50 years was obtained for all proxies as time slices for the model. In addition, all data was restricted to the window of observation of 7000--1000 BCE.

In the construction of our ‘observational model’, we considered all these proxies as informative of the number of settlements located in the north of the Swiss Alps. Consequently, we shift the causality and consideration of measurement error, which is certainly inherent in each of these proxies, to a preceding ‘process model’, in which a Poisson process describes the number of simultaneous settlements. In doing so, we establish a likelihood that estimates how credible the data is, given the model.

@@

A special class of Bayesian hierarchical models are so-called State Space Models (also known as Hidden Markov Models). These are specifically designed for time series and follow two basic principles. First, a hidden or latent process is assumed, which represents the state of the variable of interest $x_t$ through the entire time series. Over time, it is assumed that every state of variable x in the future, as well as in the past, is bound by a Markov process to the state of variable $x$ at time $t$. At the same time, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $x$ at time $t$. This means that there is a relationship between the variable $x$ and the state of variable y over time. This implies that a relationship between the individual states of variable $y$ is generated over time via the hidden variable $x$, which itself is not observable.

>>>>

  Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density. Using the  area North of the main ridge of the Swiss Alps in prehistoric times (6000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone. The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research. Our prediction of population development of the case study area accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed probability distribution of radiocarbon dates.

@@

  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (6000–1000 BCE)

@@

Prehistorians have long recognised demography as a fundamental force in human cultural evolution [@childe_man_1936]. Despite decades of interest in the demographic dynamics of prehistoric societies, concrete estimates of population size and density before written records remain elusive. Though the archaeological record provides multiple possible demographic proxies [@muller_tracing_2019], a lack of access to this data and methodological tools for turning it into quantitative estimates has left the conclusions drawn from it vague and superficial [@hassan_demographic_1981]. As a result, 'expert estimates' transferred from ethnographic parallels have often taken the place of direct inference from archaeological evidence [@morris_measure_2013, @turchin_seshat_2015].

Prehistoric demography has experienced a resurgence in recent years [@shennan_population_2000; @riede_climate_2009 and others in same issue], partly explained by a renewed interest in human–environment relations and human impact, necessarily requiring an assessment of population size. Kintigh et al. [-@kintigh_grand_2014] list human influence, dominance, population size, and population growth amongst their 'grand challenges' for archaeology in the 21st century.

In the last decade, the 'dates as data' technique [@rick_dates_1987], using the frequency of radiocarbon dates as a proxy for population dynamics, was taken up and significantly developed [e.g. @shennan_regional_2013], and has since been widely applied to archaeological contexts worldwide [@crema_review_2022]. This approach has contributed greatly to our understanding of prehistoric demography, but is not without its critics [@attenbrow_hiscock_2015, @carleton_groucutt_2020]. While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to fundamental problems common to all approaches relying on a single proxy [@french_manifesto_2021, @schmidt_approaching_2021]. We believe that these problems cannot be overcome by methodological refinements in this area alone. Instead, a Bayesian approach offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates.

@@

Proxies currently used for the estimation of population size in prehistory [following @muller_tracing_2019] can roughly be divided into three groups: ethnographic analogies; deductive estimates from ecological/economic factors; and the interpolation of frequencies of archaeological features (e.g. settlements, structures, individual finds). Three basic problems are common to all these approaches:

1.  **Reliance on single proxy**: Most investigations are using only one source of evidence. Although multi-proxy approaches exist, there the individual proxies only serve to support each other or the main estimator, without explicitly combining them to reduce the inaccuracy and biases inherent to single proxies.
2.  **Uncertainty in measurements**: All archaeological evidence is inherently uncertain which is carried through to derived measurements. However, in most studies, single curves are presented as estimates, and the potential error associated is almost never specified. 
3.  **Lack of a transfer function**: By 'transfer function', we mean something that allows for the proxy data to be interpreted in terms of actual population size or density. These could be absolute, i.e. a numerical estimate of population, or relative, i.e. a means of scaling changes in the proxy value to changes in population. Lack of suitable frameworks and 'calibration' data means that this is rarely presented alongside proxy estimates. In the best cases, there is a qualitative assessment of the informative value of the proxy, not sufficiently accounting for the complex nature of archaeological data.

Furthermore, the types of archaeological data commonly used as population proxies share a number of problematic characteristics, being:

-   **Limited**: We have only incomplete data, and they are usually not very informative.
-   **Unevenly distributed**: For example, although there is a good data on settlement frequencies for some regions, these regions are very unevenly distributed over time and space.
-   **Noisy**: Frequently individual proxies are strongly influenced by factors unrelated to population, for example taphonomic conditions or depositional biases.
-   **Unreliable**: Research strategies, research history and varying levels of resources available to researchers strongly affect the nature of compiled datasets. Systematic distortions are the rule rather than the exception.
-   **Heterogeneous**: All potential proxies have different spatio-temporal scales, granularity, information value, scales, and data formats.
-   **Indirect**: We will never have direct data on prehistoric population; only proxy data that is thought to be a reasonable substitute. The transfer functions linking the proxy data with the desired quantity (population) are unknown.
-   **Contradictory**: When considering several proxies, differences in transfer functions, data quality and noisiness inevitably lead to different results.

We argue that many, if not all, of these problems can be ameliorated through a) the explicit, quantitative integration of multiple proxies; and b) the use of a Bayesian approach to take account of and estimate uncertainty.

@@

Many of the problems with archaeological population proxies are shared with contemporary demography. In response, demographers have increasingly turned to Bayesian methods to estimate and forecast contemporary population dynamics. For example, Bryant and Zhang [-@bryant2018] consider Bayesian data modelling a solution to exactly the kind of problems that affect archaeological data. Bayesian approaches are well suited for limited, unreliable and noisy data. Various data sources, even contradictory data, can be brought into a common framework and used to support one another. These methods also provide a quantitative estimate of the likelihood and uncertainty of the model's resulting predictions (or in our case retro-dictions). Bayesian approaches are also capable of accounting for spatially and temporally incomplete data: where this data is missing, the uncertainty increases, but this does not prevent general modelling and estimation. Finally, hierarchically-structured model suites, with sub-models for each individual proxy, can be used to estimate transfer functions between them and the value to be modelled, thanks to the interaction of a large number of evidence.

This modeling technique can thus be used to join different lines of evidence horizontally and vertically and combine their results into a overall estimate, including an assessment of their reliability: contradicting data lead to a lower overall reliability, while a mutual support to smaller confidence intervals. If there is no systematic bias that affects all data sources to the same extent, this results in the most reliable estimate possible through the most heterogeneous set of data sources.

The well-established application of a Bayesian approach in radiocarbon calibration, where it is used to model radiometric uncertainty based on stratigraphic information, represents a very similar use case in archaeology. More recently, archaeologists have also used Bayesian modelling techniques for testing hypotheses relating to demographic models based on ^14^C data [e.g. @crema2021a]. This approach differs from the one presented here in that, in these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only. This is a clear step forward to a model-based, scientific analysis. However, the use of only one proxy, exclusively for testing hypotheses developed independently, creates problems comparable to those of the inductive approaches used so far: lacking a combination with other indicators, one is limited to the problems and conditions of sum calibration. Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of time series.

We attempt to make Bayesian hierarchical techniques usable for archaeological reconstructions. We want to show, in a reproducible and practical form using a case study, how Bayesian methods can make a decisive contribution to a better assessment of population development, crucial for the reconstruction of the human past, even in for periods for which we only have very patchy, noisy and unreliable data.

@@

Bayesian statistics rely on the premise that there is always some prior assumption, even if very rough, about the probability of an event. This assumption is adjusted by observing data, by checking how credible these priors are [likelihood, see also @bryant2018, 66]. This is Bayesian updating [cf. also @kruschke2015, especially 15--25], resulting in the posterior probability distribution, which represents not a point prediction. Small amounts of data lead to a broad distribution not strongly localised and restricted. Thus, we simultaneously obtained a result and an estimate the credibility interval, given the data.

This Bayesian learning is iterative and sequential, so that the result of one Bayesian inference can form the prior of another [@kruschke2015, 17]. This allows different information to be combined [@bryant2018, 219--224], as it has long been exploited by archaeology in using stratigraphic information to make radiometric dating more accurate [@ramsey1995].

This also makes a hierarchical formulation of problem domains possible. Parameters that are necessary for an estimation, such as the relationship of population density to the deforestation signal in pollen data, need not be specified explicitly, but can be given by probability distributions and then estimated in the model itself [@bryant2018, 186]. The more data available, the more degrees of freedom can be estimated with a reasonable width of credibility intervals [@kruschke2015, 112]. For the estimation of these parameters, submodels have to be created describing the relationship of the data to the characteristics of the parameter [@kruschke2015, 221--222].

@@

Our case study area north of the Swiss Alps (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland’s territory and comprises the partly flat, but largely hilly area between the Jura Mountains and the Alps. It is favourable for settlement and agriculture, especially the Swiss Plateau between Lake Zurich and Lake Geneva, by far the most densely populated region of the Switzerland today. This serves as core region in our case study, because here archaeological data is most abundant and accessible. Depending on the region, it has a very diverse natural landscape: shaped by glaciers during the ice ages, the many lakes and bogs provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, and a rich source for vegetation reconstructions by means of pollen analyses. Thanks to the very active and efficient archaeological research and heritage management there is an abundance of archaeological information, including known sites as well as dendrochronological and ^14^C data.

@@

Our case study targets the time window between 6000–1000 BCE. Its lower limit was chosen to avoid the so-called 'Hallstatt plateau' on the Northern Hemisphere radiocarbon calibration curve, causing difficulties for the ^14^C proxy. The upper limit coincides with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence.

A large number of different proxies can be integrated into an overall model, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created. Table \@ref(tab:tableproxies) provides a non-exhaustive list. For our case study, we used: a landscape openness indicator; an aoristic sum of typological dated sites; a sum calibration; and frequency data for dendro-dated lakeshore settlements in the Three Lakes region (western Swiss Plateau).

@@

From the Neolithic onwards, known settlement areas in Switzerland concentrate along its rivers and lakes [@christianlüthi2009]. Thus, our working region offers excellent data for demographic estimation, but poses very specific problems for such an undertaking. We have high-resolution information on the temporal sequence of individual lakeside settlements by means of dendro data. In these cases, ^14^C data are not as abundant simply because they are inferior to dendro dating.

@@

The dataset we use for the number of dendro-dated wetland settlements in the Three Lakes region was collected by Julian Laabs for his PhD thesis [@laabs2019]. The time series used here runs from `r abs(min(dendro$date))` to `r abs(max(dendro$date))` BCE, and contains the number of chronologically registered fell phases at individual settlements.

@@

The dataset for the ^14^C sum calibration primarily consists of data from the XRONOS database (https://xronos.ch), supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of Martínez-Grau et al. [-@martínez-grau2021]. It contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites (see Figure \@ref(fig:c14map)). The dates in the dataset range in ^14^C years from `r max(c14$bp)` to `r min(c14$bp)` uncal BP. This time window extends beyond the study horizon in order to minimise boundary effects.

We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites. For the creation of the sum calibration, the corresponding functions of the R package rcarbon [@crema2021] were used with their default settings.

@@

@@

We include relative dating information obtained from the Swiss cantonal heritage management (Figure \@ref(fig:aoristmap)). These are primarily derived from scattered surface finds, often with a low dating accuracy (only in the range of archaeological periods), incorporate into our model as a typologically dated aorist time series. It is not bound to the problems of radiocarbon dating and thus methodological issues of sum calibration can be avoided. Data from `r nrow(aorist)` sites were included in the aoristic sum.

```{r aoristcurve, fig.cap="Aoristic sum of the archaeological sites used in this analysis."}
joined_data <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "preprocessed_data", "all_proxies.csv"))
ggplot(joined_data) + geom_line(aes(x=(1950-age)*-1, y=aoristic_sum, color = "darkgray"), alpha = 0.75) +
  theme_minimal()  +  scale_x_reverse() +
    labs(x = "years BCE",
         y = "aoristic sum",
         color = "Legend") + 
    scale_color_manual(values = "darkgray", labels = "Aoristic Sum")+
  theme(legend.position="bottom")
```

@@

The natural conditions of the lakes of the Swiss Plateau enable not only highly precise dating of archaeological sites, but also a very dense network of pollen analysis. We make use of this by generating a supra-regional openness indicator for the vegetation from the pollen data (Figure \@ref(fig:pollensites)). This proxy has the specific advantage that it is not dependent on archaeological preservation conditions, making it particularly valuable for compensating systematic distortions that result from archaeological taphonomy and temporally specific settlement patterns.

@@

We assume that the higher the population density in an area, the greater the human influence on the natural environment [@zimmermann2004, @lechterbeck2014]. Evidence of deforestation can therefore provide indications of population dynamics. The full procedure for deriving this proxy from several different pollen diagrams is detailed in a previous publication [@heitz2021]. Here, we use five pollen diagrams from sites mainly in the hinterland of the large Alpine lakes.

@@

@@

The sum calibration, openness and the dendro-dated settlement data was smoothed by a moving average with a 50 years window, corresponding to the unified sampling interval for all proxies. The aoristic sum was not smoothed, because it already has a very coarse temporal resolution. In the construction of our ‘observational model’, we considered all these proxies as informative of the number of settlements located in the north of the Swiss Alps. Population development is simulated in a ‘process model’ using a Poisson process.

@@

A special class of Bayesian hierarchical models are so-called State Space Models, specifically designed for time series following two principles. First, a hidden or latent process is assumed, representing the state of the variable of interest $x_t$ through the entire time series. Every state of variable x in the future, as well as in the past, is bound by a Markov process to the state of variable $x$ at time $t$. Second, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $x$ at time $t$. This implies that a relationship between the individual states of variable $y$ is generated over time via the hidden variable $x$, which itself is not observable.

This structure makes these models particularly suitable for demographic reconstruction using archaeological and other data. The population density itself is not directly measurable. All we have at our disposal are observations derived by unknown transfer functions.

The overall model for the estimation of demographic developments is broken down into several hierarchically interconnected individual elements. The a process model represents the demographic development itself, without already being explicitly parameterised with data. Here we assume that the latent variable 'number of sites' is strongly autocorrelated across different time periods. The number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on. The population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time.
